# OZNAL - Project 01 - Classification and Regression Hypothesis

## FIIT STU - Bratislava

## March 2024

## Authors

-   Leonard Puškáč
-   Ema Richnáková

## Dataset

In this mini-project we will analyse data from <https://www.kaggle.com/datasets/julianoorlandi/spotify-top-songs-and-audio-features?resource=download>

The dataset contains information about various "top" songs and their stats on Spotify. There are 6513 entries each containing FILL IN features, such as the name of the song, its id, features that can be used to classify the songs such as the key or the mode (Major/Minor), as well as numeric features that can be used for regression analysis - energy, danceability, speechiness (the amount of spoken word), liveness, loudness, tempo and many others.

### Data Description

| Column name      | Description                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                        |
|------------------------------------|------------------------------------|
| id               | The Spotify ID for the track.                                                                                                                                                                                                                                                                                                                                                                                                                                                                                      |
| artist_names     | The name of the artist.                                                                                                                                                                                                                                                                                                                                                                                                                                                                                            |
| track_name       | The name of the track.                                                                                                                                                                                                                                                                                                                                                                                                                                                                                             |
| source           | The name of the record label.                                                                                                                                                                                                                                                                                                                                                                                                                                                                                      |
| key              | The key the track is in.                                                                                                                                                                                                                                                                                                                                                                                                                                                                                           |
| mode             | Mode indicates the modality (major or minor) of a track, the type of scale from which its melodic content is derived.                                                                                                                                                                                                                                                                                                                                                                                              |
| time_signature   | An estimated time signature. The time signature (meter) is a notational convention to specify how many beats are in each bar (or measure). The time signature ranges from 3 to 7.                                                                                                                                                                                                                                                                                                                                  |
| danceability     | Danceability describes how suitable a track is for dancing based on a combination of musical elements including tempo, rhythm stability, beat strength, and overall regularity. A value of 0.0 is least danceable and 1.0 is most danceable.                                                                                                                                                                                                                                                                       |
| energy           | Energy is a measure from 0.0 to 1.0 and represents a perceptual measure of intensity and activity. Typically, energetic tracks feel fast, loud, and noisy. For example, death metal has high energy, while a Bach prelude scores low on the scale. Perceptual features contributing to this attribute include dynamic range, perceived loudness, timbre, onset rate, and general entropy.                                                                                                                          |
| speechiness      | Speechiness detects the presence of spoken words in a track. The more exclusively speech-like the recording (e.g. talk show, audio book, poetry), the closer to 1.0 the attribute value. Values above 0.66 describe tracks that are probably made entirely of spoken words. Values between 0.33 and 0.66 describe tracks that may contain both music and speech, either in sections or layered, including such cases as rap music. Values below 0.33 most likely represent music and other non-speech-like tracks. |
| acousticness     | A confidence measure from 0.0 to 1.0 of whether the track is acoustic. 1.0 represents high confidence the track is acoustic.                                                                                                                                                                                                                                                                                                                                                                                       |
| instrumentalness | Predicts whether a track contains no vocals. "Ooh" and "aah" sounds are treated as instrumental in this context. Rap or spoken word tracks are clearly "vocal". The closer the instrumentalness value is to 1.0, the greater likelihood the track contains no vocal content. Values above 0.5 are intended to represent instrumental tracks, but confidence is higher as the value approaches 1.0.                                                                                                                 |
| liveness         | Detects the presence of an audience in the recording. Higher liveness values represent an increased probability that the track was performed live. A value above 0.8 provides strong likelihood that the track is live.                                                                                                                                                                                                                                                                                            |
| valence          | A measure from 0.0 to 1.0 describing the musical positiveness conveyed by a track. Tracks with high valence sound more positive (e.g. happy, cheerful, euphoric), while tracks with low valence sound more negative (e.g. sad, depressed, angry).                                                                                                                                                                                                                                                                  |
| loudness         | The overall loudness of a track in decibels (dB). Loudness values are averaged across the entire track and are useful for comparing relative loudness of tracks. Loudness is the quality of a sound that is the primary psychological correlate of physical strength (amplitude). Values typically range between -60 and 0 db.                                                                                                                                                                                     |
| tempo            | The overall estimated tempo of a track in beats per minute (BPM). In musical terminology, tempo is the speed or pace of a given piece and derives directly from the average beat duration.                                                                                                                                                                                                                                                                                                                         |
| duration_ms      | The duration of the track in milliseconds.                                                                                                                                                                                                                                                                                                                                                                                                                                                                         |
| weeks_on_chart   | Number of weeks the track was in the top 200 charts.                                                                                                                                                                                                                                                                                                                                                                                                                                                               |
| streams          | How many streams the track had during its period in the charts.                                                                                                                                                                                                                                                                                                                                                                                                                                                    |



## Importing Libraries

```{r}
library(magrittr)
library(tidyverse)

```

## Loading the Dataset

```{r}
data <- read_csv("spotify_top_songs_audio_features.csv", col_names = TRUE, num_threads = 4)

```

```{r}
View(data)
```

## Data Exploration

First we select the numeric features of the data set, so that we can perform correlation analysis later.

```{r}
numeric_columns <- names(data)[sapply(data, is.numeric)]
numeric_columns
```

```{r}
pair_plot_data <- data %>% select(numeric_columns)
pairs(pair_plot_data)
```

```{r}
numeric_correlation_matrix <- cor(pair_plot_data)
print(numeric_correlation_matrix)
heatmap(numeric_correlation_matrix, 
        col = colorRampPalette(c("blue", "white", "red"))(100),
        symm = TRUE)
```

### Interpreting the correlation matrix

We have selected the numeric features of the data set to see which pairs of features are worth exploring in terms of forming a hypothesis when it comes to regression. From the heatmap (and the printed numeric values) representing the correlation of these sets of numbers, we can see that there are pairs of features that clearly correlate a lot more (positive correlation relationship) than others - for example the pairs:

-   loudness - energy
-   streams - weeks_on_chart

These two pairs of features have the highest correlation coefficients among all the pairs. For our purposes we will probably be disregarding the streams an weeks_on_chart pair as forming a hypothesis about this pair is not as interesting (it seems rather clear that with more time on the chart the number of streams would increase - although it would be interesting to explore where we can expect a drop-off or what the out-liers of this trend are). We will therefore be focusing on the loudness and energy pair.

As well as that, we can see that there are certain pairs of features that have a really high negative correlation relationship (a correlation coefficient close to -1, rather than 1), meaning that as the value of one variable increases the other one tends to decreases. These pairs are:

-   acousticness - loudness
-   acousticness - energy

A high number close to either of the extremes (-1 and 1) means that there is a strong linear relationship between the variables (negative or positive) - it is therefore "safe" to make assumptions about these pairs.

However, there are still some pairs of features that might be worth exploring further in spite of their unimpressive correlation coefficients. These are features that don't correlate as much as the previously mentioned pairs, but ones which correlate (positively or negatively) strongly enough to be worth a consideration:

-   acousticness - danceability
-   valence - energy
-   valence - danceability


## Models

Based on the results of our data exploration we've come up with several hypotheses which we will test using classification and regression models. 

### Regression model

#### Hypothesis 1

*Songs tend to have higher danceability scores when they exhibit higher levels of energy.*

#### Hypothesis 2

*Songs tend to be more danceable when they have a higher tempo.*

#### Hypothesis 3

*Songs tend to have higher danceability when they exhibit lower levels of speechiness.*

### Classification model

Classification - *mode*:

-   1 - major
-   0 - minor

#### Hypothesis 1

*Songs classified as major mode are more likely to have higher danceability scores compared to songs classified as minor mode.*

#### Hypothesis 2

*Songs in the minor mode tend to have lower liveness scores.*